\documentclass[11pt,oneside]{amsbook}

\title{Introduction to linear algebra}
\author{Course notes written by Samuel Coskey\\Based on material from ``Introduction to linear algebra'' by Gilbert Strang}

\usepackage[vscale=.8,vmarginratio=4:3]{geometry}
\usepackage{mathpazo,amssymb}
\usepackage{setspace}\onehalfspacing\raggedbottom
\renewcommand{\labelitemi}{$\circ$}
\renewcommand{\labelenumi}{(\alph{enumi})}
\renewcommand{\chaptername}{Part}
\renewcommand{\thechapter}{\Roman{chapter}}
\usepackage{remreset}
\makeatletter\@removefromreset{section}{chapter}\makeatother
\usepackage{etoolbox}
\makeatletter
\pretocmd{\@seccntformat}{\S}{}{}
\patchcmd{\tocsection}{#2.}{\S#2.}{}{}
\apptocmd{\tocsection}{\dotfill}{}{}
\patchcmd{\@maketitle}{\newpage}{}{}{}
\makeatother
\usepackage[linktoc=all]{hyperref}
\usepackage{systeme}
\usepackage{tikz}

\newcommand{\set}[1]{\left\{\,#1\,\right\}}
\newcommand{\NN}{\mathbb N}
\newcommand{\ZZ}{\mathbb Z}
\newcommand{\QQ}{\mathbb Q}
\newcommand{\RR}{\mathbb R}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\rng}{rng}

\theoremstyle{definition}
\newtheorem{exerc}{Exercise}[section]
\swapnumbers
\theoremstyle{plain}
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\theoremstyle{definition}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{remark}
\newtheorem{rem}[thm]{Remark}
\newtheorem{example}[thm]{Example}
\newtheorem*{notes}{Notes and further reading}
\numberwithin{equation}{section}
\numberwithin{figure}{section}
\renewcommand{\theequation}{\arabic{section}.e\arabic{equation}}
\renewcommand{\thefigure}{\arabic{section}.f\arabic{figure}}

\begin{document}

\maketitle

\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Systems of linear equations, vectors, and matrices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Systems of linear equations and vectors}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.1.

\begin{itemize}
\item What is the equation of a line? Students may think of several answers, but the \emph{standard form} is $ax+by=c$. Why is it standard? It treats the variables symmetrically.
\item In this course we will look at generalizations of this simple type of equation
\item First we will consider the case when we have more than two variables. For example $ax+by+cz=d$. This is not the equation of a line (what is it?). But due to its similarity we still call it a \emph{linear equation}.
\item Second we will consider the case when we have more than one equation. For example, $ax+by=c$; $dx+ey=f$. This is called a \emph{system of linear equations}.
\item Let's look at several examples of simple problems which can be modeled by a system of linear equations.
\item Example 1. Economics. The NYT reported that ``The Interview'' had two million online transactions, split between rentals and downloads. A rental costs \$6 and a download costs \$15. The total revenue was \$15 million. The NYT could not determine how many transactions were rentals and how many were sales. Can you help them?
\[\systeme{r+s=2\text{ million},6r+15s=15\text{ million}}
\]
\item Example 2. Flows. Vehicles driving in each segment of a one-way grid are counted during a period of one hour. The results are shown in the figure.
  \begin{center}
    \begin{tikzpicture}[shorten <=2pt,shorten >=2pt,scale=1.5]
      \draw[->] (0,1) -- node{$520$} (1,1);
      \draw[->] (1,1) -- node{$w$} (2,1);
      \draw[->] (2,1) -- node{$600$} (3,1);

      \draw[<-] (0,2) -- node{$610$} (1,2);
      \draw[<-] (1,2) -- node{$y$} (2,2);
      \draw[<-] (2,2) -- node{$640$} (3,2);

      \draw[<-] (1,0) -- node{$480$} (1,1);
      \draw[<-] (1,1) -- node{$z$} (1,2);
      \draw[<-] (1,2) -- node{$450$} (1,3);

      \draw[->] (2,0) -- node{$390$} (2,1);
      \draw[->] (2,1) -- node{$x$} (2,2);
      \draw[->] (2,2) -- node{$310$} (2,3);
    \end{tikzpicture}
  \end{center}
  Assume that for each intersection, the number of cars entering it is the same as the number of cars leaving it. What are the values of the unknowns?
  \[\begin{cases}x+640=y+310\\y+450=z+610\\z+520=w+480\\w+390=x+600\end{cases}
  \]
\item Example 3. Trend lines. Given sampled data points, what is the equation of a trend line that best describes the overall behavior of the data? For example suppose you are measuring the change in a quantity over time. At time $t=2$ you measure $3$, at time $t=3$ you measure $5$, and at time $t=6$ you measure $6$.
  \[\systeme{49m+11b-57=0,11m+3b-14=0}
  \]
\item Before we study how to solve these types of systems, let us visualize what is going on. Consider the example system
  \[\systeme{x-2y=1,3x+2y=11}
  \]
\item The classical way to visualize this is by graphing the two lines and looking for the point where they intersect.
  \begin{center}
    \begin{tikzpicture}
      \draw[gray!30,very thin,dashed] (-0.9,-0.9) grid (4.9,4.9);
      \draw[->,thick] (-1,0)--(5,0);
      \draw[->,thick] (0,-1)--(0,5);
      \draw[domain=0:4] plot (\x,.5*\x-.5);
      \draw[domain=1:4] plot (\x,11/2-3/2*\x);
    \end{tikzpicture}
  \end{center}
\item However, another revealing way to visualize this system is to consider it as a \emph{single system} with \emph{vector coefficients}.
  \[\begin{bmatrix}1\\3\end{bmatrix}x+\begin{bmatrix}-2\\2\end{bmatrix}y
    =\begin{bmatrix}1\\11\end{bmatrix}
  \]
\item Note that we write our vectors vertically instead of horizontally, as one often does in calc 3.
\item Recall that we scale vectors by lenghening and shortening them, and that we add vectors ``tail to tip''. Thus this system is asking, how many of the vector $(1,2)$ and how many of the vector $(-2,2)$ do we need to obtain $(1,11)$?

  (figure)
\item Either way the answer is $x=3$, $y=1$.
\item Note that geogebra can be extremely helpful to visualize the row picture! It also shows the effect of elimination on the system.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Elimination}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.2.

\begin{itemize}
\item In the previous section we agued that it is important to solve systems of linear equations.
\item In this section we provide an efficient method for solving systems of linear equations: \emph{elimination}.
\item We should note that there are thousands of ways to solve systems, and no one is necessarily right or wrong. For example, many students solved for one variable in one equation, and then plugged the result into the second equation.
\item In this class you are expected to be able to solve systems using one prescribed and deterministic method, the elimination method. The benefit is that this way, all humans and computers will use the same exact steps. We will see other benefits in later material.
\item To illustrate the method, let's use a small toy system.
  \[\systeme{x-2y=1,3x+2y=11}
  \]
  \begin{itemize}
  \item Eliminate the 3, resulting in
    \[\systeme{x-2y=1,8y=8}
    \]
  \item We notate the operation by $R_2-3R_1\to R_2$.
  \item The result is called \emph{triangular} because only the upper-right entries are nonzero.
  \item The system can now be easily solved by back-substitution. The solution is $x=3$, $y=1$.
  \item The new system is equivalent to the old system, so this is the correct solution to the original problem.
  \end{itemize}
\item In the previous example, there was only one step of elimination. For larger systems there will be several steps. To illustrate the full process let us try a $3\times3$ example.
  \[\systeme{2x+4y-2z=2,4x+9y-3z=8,-2x-3y+7z=10}
  \]
  \begin{itemize}
  \item We first use the upper-left $2$ as the \emph{pivot}. That is, we use it to eliminate the coefficients below it using the operations $R_2-2R_1\to R_2$ and $R_3+1R_1\to R_3$.
  \item We then move to the next column and next row to find the new pivot. We eliminate the coefficent below it with the operation $R_3-1R_2\to R_3$.
  \item The system is upper triangular:
    \[\systeme{2x+4y-2z=2,y+z=4,4z=8}
    \]
  \item We can now back-solve starting with the last variable. The solution is $-1,2,2$.
  \end{itemize}
\item That was a case when everything went right. In fact a few strange things can happen.
\item Example requiring an exchange of equations. Note that we refuse to use a $0$ coefficient as a pivot.
  \[\systeme{5y-7z=2,2x+y+z=1,x-y+z=3}
  \]
\item Example with not enough pivots. In this case the system either has zero or infinitely many solutions, depending on whether the equations are consistent.
  \[\systeme{x-y+2z=1,2x+y-z=-1,5x-2y+5z=2}
  \]
\item Another example with not enough pivots.
  \[\systeme{x-y+2z=1,2x-2y+z=3,-x+y-z=-1}
  \]
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Matrix operations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.4.

\begin{itemize}
\item When writing systems of linear equations, we rewrite the same variable over and over again. In the future we will untangle the matrix of coefficients from the vector of variables and the vector of constants.
\item In order to understand this we need to recall the matrix-times-vector operation. Here each row of the matrix is multiplied by the vector, sum-of-products style, to obtain a new vector.
  \[\begin{bmatrix}2&3\\-1&2\end{bmatrix}
    \begin{bmatrix}5\\7\end{bmatrix}
    =\begin{bmatrix}(2)(5)+3(7)\\(-1)(5)+(2)(7)\end{bmatrix}
    =\begin{bmatrix}31\\9\end{bmatrix}
  \]
\item For example a system
  \[\systeme{3x-2y+z=4,2x+2y-z=0,-x+y+z=1}
  \]
  Would now be written
  \[\begin{bmatrix}3&-2&1\\2&2&-1\\-1&1&1\end{bmatrix}
    \begin{bmatrix}x\\y\\z\end{bmatrix}
    =\begin{bmatrix}4\\0\\1\end{bmatrix}
  \]
\item Like vectors, matrices can be scaled or added together. [Do a random example.]
\item Unlike vectors, matrices can also be multiplied together. Here each row of the first matrix is multiplied by each column of the second matrix, sum-of-products style. [Do a random square example.]
\item For nonsquare matrices, this doesn't always make sense. For it to work, the number of columns of the first matrix must match the number of rows of the second matrix. [Do a random nonsquare example.]
\item Returning to square matrices, the presence of additon and multiplication means we can do algebra. It is natural to ask whether this algebra obeys the same laws as traditional algebra: associativity, commutativity, and distributivity.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Elimination using matrices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.3.

\begin{itemize}
\item We have previously seen that a system of linear equations can be viewed as a single matrix equation $A\mathbf{x}=\mathbf{b}$. [example]
\item In this section we will see that the row elimination process is just multiplying both sides by a matrix.
\item Example
  \[\systeme{2x+4y-2z=2,4x+9y-3z=8,-2x-3y+7z=10}
  \]
  We rewrite it as
  \[\begin{bmatrix}2&4&-2\\4&9&-3\\-2&-3&7\end{bmatrix}
    \begin{bmatrix}x\\y\\z\end{bmatrix}
    =\begin{bmatrix}2\\8\\10\end{bmatrix}
  \]
\item The first step of elimination would be $R_2-2R_1\to R_2$. We can carry this out by multiplying both sides (on the left) by the ``elimination matrix''
  \[\begin{bmatrix}1&&\\-2&1&\\&&1\end{bmatrix}
  \]
  [Carry this out]
\item Every step of row elimination has a corresponding elimination matrix.
  \begin{itemize}
  \item The step $R_i+\ell R_j\to R_i$ has the matrix with $1$'s down the diagonal and an $\ell$ in position $(i,j)$.
  \item The step $R_i\leftrightarrow R_j$ has a matrix with $1$'s down the diagonal except row $i$ has a $1$ in position $j$ and row $j$ has a $1$ in position $i$.
  \item The (so far unused step) $kR_i\to R_i$ has a matrix with $1$'s down the diagonal except for a $k$ in position $(i,i)$.
  \end{itemize}
\item This means elimination can be viewed as successive multiplication by elimination matrices:
  \[E_k\cdots E_2E_1A\mathbf{x}=E_k\cdots E_2E_1\mathbf{b}
  \]
  Until the left hand side is upper triangular:
  \[U\mathbf{x}=\mathbf{b}_{\text{new}}
  \]
\item Complete the above example, keeping track of the elimination matrices.
\item Here is another example.
  \[\begin{bmatrix}2&3&1\\-2&-3&-2\\4&1&2\end{bmatrix}
    \begin{bmatrix}x\\y\\z\end{bmatrix}
    =\begin{bmatrix}6&-7&7\end{bmatrix}
  \]
\item Note that since we do the same operation to the left and right sides, we often write them both in one combined \emph{augmented matrix}. This is the final and most compact way to write a system of linear equations. 
\item Here is another example. Suppose we are given the system:
  \[\systeme{x-2y+2z=1,-3x+y+z=-1,-2x+2y+z=1}
  \]
  We can write
  \[\begin{bmatrix}
      1&-2&2&\vline&1\\
      -3&1&1&\vline&-1\\
      -2&2&1&\vline&1
    \end{bmatrix}
  \]
  and then solve.
\item Since the augmented matrix hides the variables, it is easy to forget that it represents a system of linear equations. If you are confused, try labelling each column with the corresponding variable.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LU factorization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.6.

\begin{itemize}
\item Recall what happens when you solve a system using elimination matrices. Beginning with $A\mathbf{x}=\mathbf{b}$ you left-multiply by elmination matrices $E_i$ until you reach
  \[E_k\cdots E_2E_1A\mathbf{x}=\mathbf{b}_{\mathrm{new}}
  \]
\item We know we are done when the left-hand side is upper triangular, that is, when
  \[E_k\cdots E_2E_1A=U
  \]
\item The above equation can be rewritten
  \[A=E_1^{-1}E_2^{-1}\cdots E_k^{-1}U
  \]
  where $E^{-1}$ denotes the \emph{inverse} of $E$, that is, the matrix that \emph{undoes} the effect of $E$.
\item If the elimination matrices are lower triangular (no row exchanges) then $L=E_1^{-1}E_2^{-1}\cdots E_k^{-1}$ is lower triangular, and we have written $A$ as a product of a lower and upper triangular matrices $A=LU$.
\item There are two reasons to do this. First, just as in ordinary arithmetic, it is often useful to factor something as a product of simpler things. Second, as we shall see, it turns $A\mathbf{x}=\mathbf{b}$ into $LU\mathbf{x}=\mathbf{b}$. This is a composition of two triangular systems which can each be solved by back-solving.
\item Example. Factor the matrix $A=LU$.
  \[A=\begin{bmatrix}2&4&2\\1&5&2\\4&-1&9\end{bmatrix}
  \]
  Show the elimination matrices. In each case the inverse is just negating the off-diagonal entry. Then taking the product yields a simple result. The solution is
  \[A=\begin{bmatrix}1&&\\\frac12&1&\\2&-3&1\end{bmatrix}
    \begin{bmatrix}2&4&2\\&3&1\\&&8\end{bmatrix}
  \]
  Finally check $A=LU$.
\item Next if we have a factored system $LU\mathbf{x}=\mathbf{b}$ we can use the substitution $\mathbf{c}=U\mathbf{x}$ to break it into two triangular systems: $L\mathbf{c}=\mathbf{b}$ and then $U\mathbf{x}=\mathbf{c}$. We can solve one and then the other using back/forth solving.
\item Example.
  \[\systeme{2x+4y+2z=10,x+5y+2z=11,4x-y+9z=2}
  \]
  By the work we have done previously, the system can be broken into two:
  \[\systeme{c=10,\frac12c+d=11,2c-3d+e=2}
    \qquad\text{and}\qquad
    \systeme{2x+4y+2z=c,3y+z=d,8z=e}
  \]
  The first can easily be solved forwards with solution $c=10$, $d=6$, $e=0$. We then plug these values into the second to solve backwards with solution $x=1$, $y=2$, $z=0$. Check.
\item Here is the really cool part. Suppose the matrix $A$ models a certain system, and you want to solve $A\mathbf{x}=\mathbf{b}$ over and over again for many different $\mathbf{b}$'s. If we had to perform elimination each time, we would be spending $n^3$ operations each time. If on the other hand we factor $A=LU$ once, then we can reduce every future instance to backsolving trice, and this takes only $n^2$ operations!
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Inverse matrices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 2.5.

\begin{itemize}
\item We have talked about addition and multiplication of matrices. We haven't talked about division yet. Since multiplication is not commutative, we have left and right multiplication. As a result we will have left and right division too.
\item Thus we don't use the notation $\frac1A$, but rather the multiplicative version $A^{-1}$.
\item Definition: If $A$ is any matrix, its \emph{inverse} is the matrix $A^{-1}$ such that $AA^{-1}=A^{-1}A=I$.
\item Here $I$ again means the identity matrix of $1$'s down the diagonal. It has the multiplicative identity property: $AI=IA=A$.
\item We have already seen some examples of inverse matrices with eliminations matrices. Example.
  \[A=\begin{bmatrix}1&0\\3&1\end{bmatrix},\qquad
    A^{-1}=\begin{bmatrix}1&0\\-3&1\end{bmatrix}
  \]
\item But note that other types of matrices have inverses too.
  \[A=\begin{bmatrix}2&5\\1&3\end{bmatrix},\qquad
    A^{-1}=\begin{bmatrix}3&-5\\-1&2\end{bmatrix}
  \]
  Check that the two matrices are inverses.
\item We devote the rest of the class to showing how to calculate the inverse of a matrix. Before we do recall that we can use elimination matrices to make a given matrix upper triangular:
  \[E_k\cdots E_1A=U
  \]
  This zeros out the entries below the pivots. The idea is to \emph{continue elimination} to zero out the entries above the pivots as well. We can also ensure the pivots themselves are all $1$ leaving only the identity matrix.
  \[E_m\cdots E_k\cdots E_1A=I
  \]
\item The inverse of $A$ is now observed to be the product of all the elimination matrices.
\item In order to keep a running track of the product of the elimination matrices, we can perform a side-by-side elimination:
  \begin{align*}
    [A&\mid I]\\
    [E_1A&\mid E_1]\\
      &\vdots\\
    [E_m\cdots E_1A&\mid E_m\cdots E_1]\\
    [I&\mid A^{-1}]
  \end{align*}
\item Full example.
  \[\begin{bmatrix}1&4&3\\-1&-2&0\\2&2&3\end{bmatrix}
  \]
  Don't forget to check!
\item It is important to note that not every matrix $A$ has an inverse matrix! First of all the matrix has to be square. Next, in order for the method to work, we have to have all the pivots nonzero. A square matrix with no inverse (too few pivots) is called \emph{singular}.
\item Example?
\end{itemize}


\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Column space}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 3.1.

\begin{itemize}
\item In this section we study vectors as mathematical objects.
\item The notation for the space of real vectors with $n$ components is $\RR^n$. Thus $\RR^1$ is the real line, $\RR^2$ is the euclidean plane, $\RR^3$ is real $3$-space, and so on.
\item We have seen that vectors can be scaled and added. This makes $\RR^n$ into a \emph{vector space}. A vector space is any collection of mathematical objects that may be scaled and added.
\item As an aside, other examples of vector spaces include polynomials, continuous functions, differentiable functions, matrices, and so on.
\item In this class we only care about the vector spaces $\RR^n$ and its \emph{subspaces}.
\item In order to explain what a subspace is, first note that if $\textbf{v}_1,\ldots,\textbf{v}_k$ are any vectors then a \emph{combination} of $\textbf{v}_1,\ldots,\textbf{v}_k$ is any vector that can be obtained by summing and scaling them.
\item A \emph{subspace} of $\RR^n$ is a set of vectors $S$ with the property that all combinations of vectors in $S$ also lie in $S$.
\item Though the condition may seem technical, it turns out that all subspaces of $\RR^n$ have the shape of a point, line, plane, etc, thru the origin.
\item For an example of a subspace, consider the set
  \[S_1=\set{(x,y,z)\in\RR^3\mid x+y+z=0}
  \]
  Verify that $S_1$ is a subspace by checking that scaling or adding vectors in $S_1$ are also in $S_1$. This is what it means to be a subspace. Note that the shape of this subspace is a plane.
\item For a non-example, consider the set
  \[S_2=\set{(x,y)\in\RR^2\mid 2x+y=1}
  \]
  Observe that the summation property doesn't hold.
\item The most important example of a subspace is the column space of a matrix. If $A$ is any matrix with column vectors $\mathbf{v}_1,\ldots\mathbf{v}_n$, then the \emph{column space} of $A$ consists of all combinations $\mathbf{v}_1x_1+\cdots+\mathbf{v}_nx_n$ of the columns of $A$.
\item Equivalently, the column space of $A$ consists of all vectors $\mathbf{b}$ such that $A\mathbf{x}=\mathbf{b}$ has a solution.
\item Example. Consider the matrix
  \[A=\begin{bmatrix}1&0\\4&3\\2&3\end{bmatrix}
  \]
  The column space of $A$ consists of the plane spanned by the its two column vectors. We can test whether the vector $(-1,-1,1)$ lies in the column space of $A$ by deciding whether the system is consistent:
  \[A\mathbf{x}=\begin{bmatrix}-1\\-1\\1\end{bmatrix}
  \]
  Draw a sketchy picture showing the concept.
\item We can find an algebraic equation which governs which $\mathbf{b}$'s work. We ask when the following is consistent.
  \[\begin{bmatrix}
      1&0&\vline&b_1\\
      4&3&\vline&b_2\\
      2&3&\vline&b_3
    \end{bmatrix}
  \]
  Solve for the $b$'s that work.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Null space: the complete solution to $A\mathbf{x}=\mathbf{0}$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 3.2.

\begin{itemize}
\item In the last section we talked about the column space of a matrix $A$. We defined it as all the combinations of the columns of $A$ (the linear \emph{span} of the columns of $A$).
\item The column space is also the set of all $\mathbf{b}$ such that $A\mathbf{x}=\mathbf{b}$ has a solution. (Think range.)
\item However, as indicated on the cover of the textbook, the column space is not the only subspace we can associate with a matrix. There are three more, and in this class we investigate the second most important one, the null space.
\item Given a matrix $A$, the \emph{null space} of $A$ is the set of all $\mathbf{x}$ such that $A\mathbf{x}=\mathbf{0}$. (Think zeros or roots.)
\item Note that this is a subspace! For example if we have $A\mathbf{x^1}=\mathbf{0}$ and $A\mathbf{x^2}=\mathbf{0}$, then we have $A(\mathbf{x^1}+\mathbf{x^2})=\mathbf{0}$ too.
\item Example.
  \[A=\begin{bmatrix}
      1&2&2\\2&4&8\\3&6&10
    \end{bmatrix}
  \]
  To write the solution, we identify the variables whose column has no pivot as \emph{free variables}, and the variables whose column has a pivot as \emph{basic variables}. We solve for the basic variables \emph{in terms of} the free variables.
  \[\begin{bmatrix}x\\y\\z\end{bmatrix}
    =\begin{bmatrix}-2y\\y\\0\end{bmatrix}
  \]
\item We achieve some additional clarity if we eliminate above and below the pivots, and eliminate the pivots to $1$s. This is called the \emph{reduced row echelon form} (RREF).
\item Example.
  \[A=\begin{bmatrix}
      1&2&2&4\\3&8&6&16
    \end{bmatrix}
  \]
  We again solve for the basic variables as functions of the free variables. In fact by factoring out the free variables, we can also write the solution in \emph{parametric vector form}. The resulting numeric vectors are called the \emph{special solutions} and the free variables are the parameters.
  
  \[\begin{bmatrix}x\\y\\z\\w\end{bmatrix}
    =\begin{bmatrix}-2\\0\\1\\0\end{bmatrix}z
    +\begin{bmatrix}0\\-2\\0\\1\end{bmatrix}w
  \]
\item Since the parameters are in some sense unnecessary information, we often write just the \emph{null space matrix} which is just the matrix of special solutions.
  \[N=\begin{bmatrix}
      -2&0\\0&-2\\1&0\\0&1
    \end{bmatrix}
  \]
  It is called the null space matrix because $AN=$ the zero matrix.
\item The special solutions also provide a recipe for a combination of the columns of $A$ that gives $\mathbf{0}$.
\item Larger example if time.
  \[A=\begin{bmatrix}
      1&2&2&3\\2&4&8&10\\3&6&10&13
    \end{bmatrix}
  \]
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The complete solution to $A\mathbf{x}=\mathbf{b}$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 3.3.

\begin{itemize}
\item In the previous section we showed how to find the complete solution to a \emph{homogeneous} system, that is, a system of the form $A\mathbf{x}=\mathbf{0}$.
\item We found that the solution to such a system was a subspace and could be expressed as the combinations of the special solutions.
\item We will now use this as a launching point to solve a system of the form $A\mathbf{x}=\mathbf{b}$. It turns out that solutions to this equation all lie in a \emph{translate} of the null space.
\item To get the idea, let's look at the case of one equation in two variables. For example, first consider the homogeneous equation:
  \[2x+3y=0
  \]
  We can plot this, it is just a line through the origin with slope $-2/3$.

  Now let us consider an inhomogeneous equation with the same left-hand side:
  \[2x+3y=3
  \]
  We can plot this, it is just the above line shifted to the right by $3/2$ (or up by $1$).
\item In the matrix notation we have been using, we are looking at $A=\begin{bmatrix}2&3\end{bmatrix}$. We find that the null space of $A$ is
  \[\mathbf{x}_{\mathrm{null}}
    =\begin{bmatrix}-3/2\\1\end{bmatrix}y
  \]
  which is the line through the origin found above.

  The general solution is
  \[\mathbf{x}_{\mathrm{general}}
    =\begin{bmatrix}3/2\\0\end{bmatrix}
    +\begin{bmatrix}-3/2\\1\end{bmatrix}y
  \]
  which is the line translated away from the origin found above.
\item In general, we will find that the general solution to a system of linear equations $A\mathbf{x}=\mathbf{b}$ will be equal to a single vector (the particular solution) added to the vectors in the null space (the null or homogeneous solution). This strongly mirrors the method from systems of linear differential equations.
\item Example
  \[\begin{bmatrix}
      1&3&0&2\\0&0&1&4\\1&3&1&6
    \end{bmatrix}
    \mathbf{x}=
    \begin{bmatrix}
      1\\6\\7
    \end{bmatrix}
  \]
  \begin{itemize}
  \item Step one. Eliminate the full system to RREF.
    \[\begin{bmatrix}
        1&3&0&2&\vline&1\\
        &&1&4&\vline&6\\
        &&&&\vline&
      \end{bmatrix}
    \]
  \item Step two. Find the null space. Then forget about the right-hand side briefly and find the null space of $A$ in parametric vector form as usual.
  \item Step three. Find the particular solution to the original system by setting the free variables to zero. Note there is also a ``trick'' here using the right-hand side of RREF.
  \item Step four. The answer is the sum of the particular and general solutions.
    \[\mathbf{x}_{\mathrm{general}}
      =\begin{bmatrix}1\\0\\6\\0\end{bmatrix}
      +\begin{bmatrix}-3\\1\\0\\0\end{bmatrix}x_2
      +\begin{bmatrix}-2\\0\\-4\\1\end{bmatrix}x_4
    \]
  \end{itemize}
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Subspaces, orthogonality, and applications}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Independence and basis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 3.4.

\begin{itemize}
\item Recall that the column space of a matrix $A$ consists of all combinations of the columns of $A$. Sometimes we say the columns are a \emph{spanning set} for the column space. The idea is that to tell you about a plane, I only need to tell you two vectors on the plane, a spanning set, and not every vector.
\item But sometimes not all columns are necessary to form a spanning set. Example.
  \[A=\begin{bmatrix}1&1&3\\2&0&2\\3&1&5\end{bmatrix}
  \]
  Notice that the third column of $A$ is a combination of the first two. The first two columns already form a spanning set, making the third column in some sense redundant.
\item Geometrically, the three columns of $A$ point in distinct directions, but still don't manage to break outside of a plane.
\item We say that a set of vectors is \emph{linearly dependent} if one of the vectors is a combination of the others. Otherwise we say the set is \emph{linearly independent}.
\item Given a matrix $A$, we wish to remove the redundant columns to find a linearly independent spanning set for the column space. A linearly independent spanning set is called a \emph{basis}.
\item The basis has ``just the right number'' of vectors. A basis for a plane will have two vectors, for a three-d thingy will have three vectors, etc. The number of vectors in a basis is called the \emph{dimension} of the space.
\item Geometrically, the matrix above consists of three vectors pointing in three directions but which do not escape from a plane.
\item Given a matrix $A$, there is a simple rule to find out which columns belong in the basis and which columns are redundant: The basic (non-free) variable columns belong in the basis and the free variable columns are redundant.
\item Example.
  \[A=\begin{bmatrix}
      1&2&0&3\\1&2&-1&-1\\2&4&-1&2
    \end{bmatrix}
  \]
  The RREF form shows us exactly why the free variable columns are redundant. The entries in the column show how to express it as a combination of earlier columns.
\item We also wish to find a basis for the null space of $A$. In fact we have already done this: the special solutions are always linearly independent.
\item Do it for the example above.
\item To summarize: To find a basis for the column space of $A$, we eliminate to REF, identify the pivot columns, and selecto those columns from the original matrix $A$. To find a basis for the null space of $A$, we find the special solutions or null space matrix.
\item We can apply the same reasoning to remove the redundant elements from any set of vectors, and find a linearly independent subset with the same span.
\item Example. Find a basis for the subspace of $\RR^4$ spanned by the vectors. What is the dimension of the subspace?
\item Example.
  \[\begin{bmatrix}1\\0\\1\end{bmatrix},\quad
    \begin{bmatrix}2\\0\\2\end{bmatrix},\quad
    \begin{bmatrix}3\\1\\3\end{bmatrix},\quad
    \begin{bmatrix}2\\1\\2\end{bmatrix}
  \]
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The fundamental subspaces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 3.5.

\begin{itemize}
\item We have previously discussed two ``fundamental'' subspaces associated with a given matrix $A$: the column space and the null space.
\item The column space is the span of the columns, the null space is the combinations of the columns that give zero. It should not surprise us that there are two corresponding spaces for the rows.
\item We can study rows as columns using the transpose:
\item Given a matrix $A$ we form the \emph{transpose} matrix $A^T$ as follows: The rows of $A^T$ are the columns of $A$ and the columns of $A^T$ are the rows of $A$.
\item Example.
\item The four fundamental subspaces corresponding to a matrix $A$:
  \begin{itemize}
  \item The column space $C(A)$ is the set of all combinations of (span of) the column vectors of $A$
  \item The null space $N(A)$ is the set of all solutions to $A\mathbf{x}=\mathbf{0}$, that is, the null combinations of the columns
  \item The row space $R(A)$ is the set of all combinations of (span of) the row vectors of $A$. This is also $C(A^T)$.
  \item The left-null space $LN(A)$ is the set of all solutions to $\mathbf{y}^TA=\mathbf{0}$, that is, the null combinations of the rows. This is also $N(A^T)$.
  \end{itemize}
\item We wish to find bases and dimensions of the four fundamental subspaces corresponding to $A$. We already know how to find a basis for the column and null space.
\item To find a basis for the row space, simply take the \emph{nonzero rows} of the REF or RREF matrix.
\item To find a basis for the left-null space, there are two tough options: (1) find the null space of $A^T$; (2) keep track of elimination steps to write $EA=R$ and take the rows of $E$ corresponding to zero rows of $R$
\item There is also a trick to find a basis for the row space: it is the nonzero rows of the REF of $A$.
\item Example. Find bases and dimensions of the four fundamental subspaces.
  \[A=\begin{bmatrix}
      1&3&0&5\\2&6&1&16\\5&15&0&25
    \end{bmatrix}
  \]
  The RREF form and the elimination matrix are
  \[R=\begin{bmatrix}
      1&3&0&5\\&&1&6\\&
    \end{bmatrix},\qquad
    E=\begin{bmatrix}
      1&&\\-2&1&\\-5&&1
    \end{bmatrix}
  \]
\item The dimensions of the four fundamental subspaces can all be calculated from the shape $m\times n$ of $A$ and the number of pivots $r$.
  \begin{itemize}
  \item $C(A)$ is a $r$-dimensional subspace of $\RR^m$
  \item $N(A)$ is an $n-r$-dimensional subspace of $\RR^n$
  \item $R(A)$ is an $r$-dimensional subspace of $\RR^n$
  \item $LN(A)$ is an $m-r$-dimensional subspace of $\RR^m$
  \end{itemize}
\item The number of pivots $r$ is called the \emph{rank} of the matrix $A$ and measures the size of the matrix in terms of the dimension of its row or column space.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Orthogonality and fundamental subspaces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 4.1.

\begin{itemize}
\item We have introduced four fundamental subspaces corresponding to a matrix, and seen how at least the column space and null space are connected with solving linear systems.
\item But there is one more special property of the four subspaces that we have yet to see.
\item The term \emph{orthogonal}, when applied to vectors or subspaces, means perpendicular.
\item You may recall that two vectors $\mathbf{v},\mathbf{w}$ are orthogonal precisely when their dot product $\mathbf{v}^t\mathbf{w}=0$. We use the symbol $\mathbf{v}\perp\mathbf{w}$ do denote this.
\item Why is this the case? Briefly, it is the pythagorean theorem.
  \begin{itemize}
  \item Two vectors $\mathbf{v},\mathbf{w}$ are orthogonal exactly when the triangle they form is a right triangle.
  \item In this case we know $|\mathbf{v}|^2+|\mathbf{w}|^2=|\mathbf{v-w}|^2$.
  \item Using the distance formula (another version of the pythagorean theorem) we have $\sum v_i^2+\sum w_i^2=\sum(v_i-w_i)^2$.
  \item Distributing the terms of the right hand side, we get $\sum v_i^2+\sum w_i^2=\sum v_i^2-\sum 2v_iw_i+\sum w_i^2$.
  \item Cancelling this simply says that $\mathbf{v}^T\mathbf{w}=0$.
  \end{itemize}
\item If $V$ is a subspace of some Euclidean space $\RR^n$, its \emph{orthogonal complement} $V^\perp$ is the subspace consisting of all vectors $\mathbf{w}$ that are orthogonal to all vectors in $V$.
\item Example. If $V$ is the line in $\RR^3$ that passes through the vector $(1,-1,2)^T$, then the orthogonal complement of $V$ would be the vectors $(x,y,z)$ such that
  \[\begin{bmatrix}1&-1&2\end{bmatrix}
    \begin{bmatrix}x\\y\\z\end{bmatrix}=0
  \]
  In other words, the orthogonal compleement is the set of solutions to $x-y+2z=0$, or in other words, the plane parameterized by
  \[\begin{bmatrix}1\\1\\0\end{bmatrix}y
    +\begin{bmatrix}-2\\0\\1\end{bmatrix}z
  \]
\item It makes sense that the orthogonal complement of a line in $\RR^3$ would be a plane---the dimensions should add to $3$.
\item We saw from the example that if $A=\begin{bmatrix}1&-1&2\end{bmatrix}$ then the row space of $A$ is the orthogonal compelement of the null space of $A$.
\item Strang calls this the \emph{Fundamental theorem of linear algebra}: Given any matrix $A$, the row space and null space are orthogonal complements of each other, and the column space and left-null space are orthogonal compelements of each other.
\item We can use this property to calculate orthogonal complements.
\item Example. Find the orthogonal complement of the subspace of $\RR^4$ with the basis:
  \[\begin{bmatrix}1\\3\\0\\1\end{bmatrix},\quad
    \begin{bmatrix}1\\1\\2\\1\end{bmatrix}
  \]
  
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Orthogonal projections}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 4.2.

\begin{itemize}
\item We know how to completely solve a system $A\mathbf{x}=\mathbf{b}$ if it is consistent.
\item What if it isn't consistent, but you instead want to find the nearest possible solution $\mathbf{x}$?
\item Geometrically, we are saying that $\mathbf{b}$ doesn't lie in the column space of $A$, but we want to find the closest possible vector $\mathbf{p}$ and solve $A\mathbf{x}=\mathbf{p}$ instead.
\item From a picture, we see that an optimal solution is achieved exactly when the error $\mathbf{b}-A\hat{\mathbf{x}}$ is orthogonal to the column space of $A$.
\item The method is known as \emph{orthogonal projection}.
  \begin{itemize}
  \item Given an inconsistent equation $A\mathbf{x}=\mathbf{b}$, what vector $\hat{\mathbf{x}}$ makes $A\hat{\mathbf{x}}$ closest to $\mathbf{b}$?
  \item We note that an optimal solution is found when $C(A)$ is orthogonal to the error term $\mathbf{b}-A\hat{\mathbf{x}}$. In equations, we set
    \[A^T(\mathbf{b}-A\hat{\mathbf{x}})=0
    \]
  \item This always simplifies down to
    \[A^TA\hat{\mathbf{x}}=A^T\mathbf{b}
    \]
  \item We solve this equation to find the optimal solution $\hat{\mathbf{x}}$.
  \item We can also find the \emph{projection} $\mathbf{p}=A\hat{\mathbf{x}}$, that is, the closest vector to $\mathbf{b}$ in the column space of $A$.
  \end{itemize}
\item Example. Find the best solution to the inconsistent system
  \[\begin{bmatrix}1&0\\1&1\\1&2\end{bmatrix}\mathbf{x}
    =\begin{bmatrix}6\\0\\0\end{bmatrix}
  \]
  Step 0, 1, 2, etc.
\item The situation is somewhat simpler when $A$ is just a single vector $\mathbf{a}$. Here, we are projecting $\mathbf{b}$ onto the line through $\mathbf{a}$.
  \begin{itemize}
  \item Given a vector $\mathbf{a}$ and a vector $\mathbf{b}$, what multiple $\hat x$ makes $\mathbf{a}\hat x$ closest to $\mathbf{b}$?
  \item Once again, we find an optimal solution when $\mathbf{b}-\mathbf{a}\hat x$ is orthogonal to $\mathbf{a}$. In equations this means
    \[\mathbf{a}^T(\mathbf{b}-\mathbf{a}\hat x)=0
    \]
  \item We can actually solve this explicitly to obtain
    \[\hat x=\mathbf{a}^T\mathbf{b}/\mathbf{a}^T\mathbf{a}
    \]
  \item Once again the projection is simply
    \[\mathbf{p}=
      \frac{\mathbf{a}^T\mathbf{b}}{\mathbf{a}^T\mathbf{a}}
      \mathbf{a}
    \]
  \end{itemize}
\item Example. $\mathbf{a}=(1,2,3)$ and $\mathbf{b}=(4,6,4)$.
\item Given a matrix $A$, we can always find the \emph{projection matrix} $P$ which maps vectors $\mathbf{b}$ to their projections $\mathbf{p}$ in the column space of $A$.
\item Recalling that $A^TA\hat{\mathbf{x}}=A^T\mathbf{b}$ and $\mathbf{p}=A\hat{\mathbf{x}}$, we find that
  \[P=A(A^TA)^{-1}A^T
  \]
\item Example.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linear modeling}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 4.3.

\begin{itemize}
\item This is the same as least squares optimization in statistics. The idea is that given a sequence of data points, we want to find the line which best fits the data.
\item One can also use the method to fit more interesting curves to the data, as we will mention briefly at the end.
\item We begin with best line fitting. Given a sequence of data points $(t_1,b_1),\ldots,(t_n,b_n)$, we will try to fit them to the linear model $C+Dt=y$.
\item The idea is that the data points all lie in a line exactly when the system is consistent:
  \[\{1C+t_iD=b_i
  \]
\item Typically the system will be inconsistent, so we instead find the best approximation as before.
\item First day example.
  \begin{itemize}
  \item Find the line that best fits the data.
    \[\begin{bmatrix}0\\6\end{bmatrix}
      \begin{bmatrix}1\\0\end{bmatrix}
      \begin{bmatrix}2\\0\end{bmatrix}
    \]
  \item We first identify the system that tries to fit these to a line.
    \[\systeme{C+0D=6,C+1D=0,C+2D=0}
    \]
  \item Writing this as a matrix equation we have
    \[\begin{bmatrix}1&0\\1&1\\1&2\end{bmatrix}
      \mathbf{C}=\begin{bmatrix}6\\0\\0\end{bmatrix}
    \]
  \item Finding the best approximation we obtain $\hat C=5$ and $\hat D=-3$. Thus the best fit line is $5-3t$. Note that this sort of makes sense with the data.
  \item By the way, the projection is $(5,2,-1)$ which corresponds to the points along the best fit line. The error is $(1,-2,1)$ and corresponds to the residuals.
  \end{itemize}
\item For a larger number of data points, the $A$ matrix and $\mathbf{b}$ vector will be as below:
  \[A=\begin{bmatrix}
      1&t_1\\
      2&t_2\\
      \vdots&\vdots\\
      1&t_n
    \end{bmatrix}
    ,\quad
    \mathbf{b}=\begin{bmatrix}b_1\\b_2\\\vdots\\b_n\end{bmatrix}
  \]
\item Here is the really cool thing. The method can also be used to fit any ``linear'' model (meaning linear combination of functions).
\item For example we can find the best fitting parabola using the model $C+Dt+Et^2=y$.
  \begin{itemize}
  \item Find the parabola that best fits the data.
    \[\begin{bmatrix}0\\6\end{bmatrix}
      \begin{bmatrix}1\\0\end{bmatrix}
      \begin{bmatrix}2\\0\end{bmatrix}
      \begin{bmatrix}3\\2\end{bmatrix}
    \]
  \item This time the system we want to find our best solution is:
    \[\systeme[CDE]{1C+0D+0E=6,
        1C+1D+1E=0,1C+2D+4E=0,1C+3D+9E=2}
    \]
  \item In other words, the thing is
    \[A=\begin{bmatrix}
        1&t_1&t_1^2\\
        1&t_2&t_2^2\\
        \vdots&\vdots&\vdots\\
        1&t_n&t_n^2
      \end{bmatrix}
      ,\quad
      \mathbf{b}=\begin{bmatrix}
        b_1\\b_2\\\vdots\\b_n
      \end{bmatrix}
    \]
  \item We won't solve it, but the outcome is $y=5.8-7.2t+2t^2$.
  \end{itemize}
\item We can summarize the above theory as follows. Given fixed $t_1,\ldots,t_n$, the set of possible time series $\mathbf b=(b_1,\ldots,b_n)$ form a vector space $\RR^n$. Only some of these vectors fit the desired model, and the set of those that do forms a subspace of $\RR^n$. For any other vector $\mathbf b$, we simply project it onto that subspace to find the closest vector to it that does fit the model.
\item This is a beautiful and typical example of an application of abstract mathematics. A given element of some space ($\RR^n$) doesn't have a certain property (lie on a line), so we simply project it onto the subspace consisting of the things that do have the property.
\end{itemize}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Determinants, eigenvalues, eigenvectors, and applications}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The determinant of a matrix}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 5.1.

\begin{itemize}
\item Consider the process of inverting a generic $2\times2$ matrix
  \[A=\begin{bmatrix}a&b\\c&d\end{bmatrix}
  \]
\item Notice the last pivot is $ad-bc$, and that this quantity shows up in the denominator of the entries of $A^{-1}$. Notice also that if this quantity is zero, then the method fails and $A$ is not invertible.
\item The special quantity $ad-bc$ is called the \emph{determinant} of $A$ or $\det(A)$ or $|A|$ because it determines whether $A$ is invertible or not.
\item Class activity. Find the last pivot of the general $3\times3$ matrix.
  \[\begin{bmatrix}a&b&c\\d&e&f\\g&h&i\end{bmatrix}
  \]
\item These small cases leave it very unclear: what makes the determinant the determinant? Using the $2\times2$ and $3\times3$ determinants as motivation, we now proceed to define the determinant by its basic properties.
  \begin{itemize}
  \item (Scaling) Scaling a row of by a scalar $t$ also scales the determinant by $t$.
    \[\det\begin{bmatrix}ta&tb\\c&d\end{bmatrix}
      =tad-tbc=t(ad-bc)
    \]
  \item (Row exchange) Exchanging two rows negates the determinant.
    \[\det\begin{bmatrix}c&d\\a&b\end{bmatrix}
      =cb-da=-(ad-bc)
    \]
  \item (Elimination) Adding a of one row to another row has no effect on the determinant.
    \[\det\begin{bmatrix}a+tc&b+td\\c&d\end{bmatrix}
      =ad-bc
    \]
  \item (Identity) The determinant of the identity matrix is $1$. This is clear from small examples.
  \end{itemize}
\item Taken together, these four rules allow you to calculate the derivative of any matrix. To see how, we reason as follows.
  \begin{itemize}
  \item The determinant of a diagonal matrix is the product of its diagonal entries. To see this we use the Scaling rule repeatedly plus the Identity rule.
    \[\det\begin{bmatrix}d_1&&\\&d_2&\\&&\ddots\end{bmatrix}
      =d_1\det\begin{bmatrix}1&&\\&d_2&\\&&\ddots\end{bmatrix}
      =\cdots=d_1d_2\cdots d_n
    \]
  \item The determinant of an upper triangular matrix is the product of its diagonal entries too. To see this note that if the determinant is not zero, it has all its pivots and you can use elimination steps to make it diagonal. Then apply the previous rule.
  \item The determinant of any matrix is the product of its pivots.
  \end{itemize}
\item In conclusion, we can always find the determinant by eliminating it to upper triangular, and multiplying the pivots.
\item Warning: if you use any Scaling or Row Exchange steps while eliminating, you must keep track of this along the way.
\item Example.
  \[\det\begin{bmatrix}1&2&3\\3&0&1\\1&2&1\end{bmatrix}
  \]
\item Example.
  \[\det\begin{bmatrix}2&4&6\\2&4&1\\1&4&2\end{bmatrix}
  \]
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Formulas for the determinant}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 5.2.

\begin{itemize}
\item Last week we defined the determinant by its four key properties, and showed how to calculate determinants using elimination.
\item It is natural to ask whether there is an \emph{absolute} formula for the determinant. There is one, but we won't dwell on it because it isn't too useful.
\item To see what it is, recall the six-term formula for the determinant of a $3\times3$ matrix:
  \[\det\begin{bmatrix}a&b&c\\d&e&f\\g&h&i\end{bmatrix}
    =aei+bfg+cdh-afh-bdi-ceg
  \]
\item Notice that each term contains one entry from every row, with every column used exactly once. A selection of one entry from every row, with every column used exactly once, is called a \emph{permutaion}. A permutation is a bijective function $\sigma\colon n\to n$. The set of all permutations is denoted $S_n$. Every permutation has a \emph{sign}, which is $\pm1$ depending on whether an even or odd number of swaps are needed to reduce it to the identity.
\item The determinant is always equal to
  \[\det(A)=
    \sum_{\sigma\in S_n}\mathrm{sign}(\sigma)\prod_1^na_{i,\sigma(i)}
  \]
\item This formula is called the ``big'' formula because it has $n!$ many terms, generally considered an intractable quantity. It can occasionally be useful for sparse matrices.
\item Example:
  \[\det\begin{bmatrix}&1&&\\1&&1\\&1&&1\\&&1\end{bmatrix}
  \]
\item \textbf{Note}: Next time it would be better to use the matrix version of permutations, and to decompose the determinant as
  \[\det\begin{bmatrix}a&b&c\\d&e&f\\g&h&i\end{bmatrix}
    =\begin{bmatrix}a\\&e\\&&i\\\end{bmatrix}
    +\begin{bmatrix}&b\\&&f\\g\\\end{bmatrix}
    +\begin{bmatrix}&&c\\d\\&h\\\end{bmatrix}
    -\begin{bmatrix}a\\&&f\\&h\\\end{bmatrix}
    -\begin{bmatrix}&b\\d\\&&i\\\end{bmatrix}
    -\begin{bmatrix}&&c\\&e\\g\\\end{bmatrix}
  \]
\item In addition to the algorithmic definiton and the big formula, we finally have a recursive formula called the cofactor formula.
\item To understand this formula, we again look at the six term formula for the determinant of a $3\times3$ matrix. This time we factor some of the terms:
  \[\det\begin{bmatrix}a&b&c\\d&e&f\\g&h&i\end{bmatrix}
    =a(ei-fh)+b(fg-di)+c(dh-eg)
  \]
\item The last result is a combination of three sub-determinants or ``cofactors'':
  \[\det\begin{bmatrix}a&b&c\\d&e&f\\g&h&i\end{bmatrix}
    =a\det\begin{bmatrix}e&f\\h&i\end{bmatrix}
    -b\det\begin{bmatrix}d&f\\g&i\end{bmatrix}
    +c\det\begin{bmatrix}d&e\\g&h\end{bmatrix}
  \]
\item Given a matrix $A$ consisting of entries $a_{ij}$, we define the \emph{cofactors} $C_{ij}$ to be $(-1)^{i+j}$ times the determinant of the matrix obtained from $A$ by deleting the $i$th row and $j$th column.
\item The \emph{cofactor formula} for the determinant states that
  \[\det(A)=a_{11}C_{11}+\cdots+a_{1n}C_{1n}
  \]
\item To remember the signs of the cofactors $(-1)^{i+j}$ we use the memory device
  \[\begin{bmatrix}
      +&-&+\\-&+&+&\cdots\\+&-&+\\&\vdots&&\ddots
    \end{bmatrix}
  \]
\item Example.
  \[\det\begin{bmatrix}
      2&-1\\-1&2&-1\\&-1&2&-1\\&&-1&2
    \end{bmatrix}
  \]
\item Note that you don't have to use the top row. You can use any row or column, provided you are careful to get the signs of the cofactors just right.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two applications of determinants}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 5.3.

\begin{itemize}
\item In the next chapter, we will use determinants frequently to decide when a matrix is singular. Before moving on, we give two additional applications of determinants.
\item The first application involves the close connection between cofactors and entries in the inverse matrix.
\item The first hint of this was the formula for a $2\times 2$ matrix, where the entries are signed cofactors:
  \[\begin{bmatrix}a&b\\c&d\end{bmatrix}^{-1}
    =\frac{1}{ad-bc}\begin{bmatrix}d&-b\\-c&a\end{bmatrix}
  \]
\item Given $A$, recall that we can form the matrix $C$ consisting of cofactors.
  \[A=\begin{bmatrix}1&2&3\\4&2&2\\1&1&1\end{bmatrix},\qquad
    C=\begin{bmatrix}0&-2&2\\1&-2&1\\-2&10&-6\end{bmatrix}
  \]
\item Now try multiplying $AC^T$. What do you get?
\item The product $AC^T$ always turns out to be a scalar matrix whose diagonal entries are the determinant of $A$.
  \[\begin{bmatrix}
      a_{11}&\cdots&a_{1n}\\
      \vdots&&\vdots\\
      a_{n1}&\cdots&a_{nn}
    \end{bmatrix}
    \begin{bmatrix}
      c_{11}&\cdots&c_{1n}\\
      \vdots&&\vdots\\
      c_{n1}&\cdots&c_{nn}
    \end{bmatrix}
    =\begin{bmatrix}
      \det(A)\\&\ddots\\&&\det(A)
    \end{bmatrix}
    \]
\item While this may seem mysterious at first, this really just comes from the cofactor expansion of the determinant. Every diagonal entry of $AC^T$ is of the form $a_{i1}c_{i1}+\cdots+a_{in}c_{in}$ and this equals $\det(A)$.
\item Note that the off-diagonal entries are all zero because they correspond to the cofactor expansion of the determinant of a matrix with a repeated row.
\item In shorter terms, the formula says that
  \[AC^T=\det(A)I
  \]
\item Thus we arrive at Cramer's rule for the inverse of a matrix, which states that
  \[A^{-1}=\frac{1}{\det(A)}C^T
  \]
\item Example.
  \[A=\begin{bmatrix}1\\1&1\\1&1&1\\1&1&1&1\end{bmatrix},\quad
    A^{-1}=\begin{bmatrix}1\\-1&1\\&-1&1\\&&-1&1\end{bmatrix}
  \]
\item The second application is about the calculation of the area of a parallelepiped. Recall that a parallelepiped is the higher dimensional analog of a paralellogram and is the a bound by several vectors and their sums.
\item Let's begin with the $2$-dimensional parallelogram with vector sides $(a,b)$ and $(c,d)$.
  \begin{center}
    \begin{tikzpicture}[yscale=.7]
      \draw[->] (-.5,0)--(5.5,0);
      \draw[->] (0,-.5)--(0,3.5);
      \draw (0,0) rectangle (5,3);
      \draw (0,0) -- (1,2) -- (5,3) -- (4,1) -- (0,0);
      \draw (4,0) -- (4,1) -- (5,1);
      \draw (0,2) -- (1,2) -- (1,3);
      \node[below] at (2,0) {$a$};
      \node[right] at (5,.5) {$b$};
      \node[below] at (4.5,0) {$c$};
      \node[right] at (5,2) {$d$};
    \end{tikzpicture}
  \end{center}
\item We find that the area of the paralellogram itself is $ad-bc$, the determinant of the matrix consisting of the two vector sides.
\item This turns out to be true in higher dimensions, so that the $n$-dimensional volume of the paralellepiped with vector sides $\mathbf{v}_1,\ldots,\mathbf{v}_n$ is given by
  \[V=\left|\det\begin{bmatrix}
        |&&|\\
        \mathbf{v}_1&\cdots&\mathbf{v}_n\\
        |&&|
      \end{bmatrix}\right|
  \]
\item To understand briefly how this can be true, we can see that the area of a parallelepiped satisfies the same properties 1--4 that we used to define the determinant.
  \begin{itemize}
  \item (Identity) If $\mathbf{v}_i=\mathbf{e}_i$ then the parallelepiped is the unit hypercube which clearly has volume $1$.
  \item (Scaling) If you scale one of the sides of a paralellepiped by $t$ this clearly scales the volume by $t$.
  \item (Elimination) Adding a multiple of one vector to another slides the parallelepiped along one of its sides. This has no effect on the volume.
  \end{itemize}
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Eigenvalues and eigenvectors}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 6.1.

\begin{itemize}
\item In the final chapter of our course, we take you through the theory and several applications of eigenvalues and eigenvectors. This is probably the most powerful and broadly used concept in linear algebra.
\item We begin with an illustrative example, which we will return to in a future session.
\item Imagine a large island with two cities, $C$ and $D$. The two cities start with equal population of $50$ thousand. Every year 20\% of the population of city $C$ moves to city $D$, and 30\% of city $D$ moves to city $C$.
  \begin{center}
    \begin{tikzpicture}
      \node at (0,0) (A) {$C$};
      \node at (2,0) (B) {$D$};
      \draw[->,bend right] (A) edge node[below]{20\%} (B);
      \draw[->,bend right] (B) edge node[above]{30\%} (A);
    \end{tikzpicture}
  \end{center}
\item What happens to the population over time? Any guesses? To get an idea, let's look at the first few years.
  \begin{center}
    \begin{tabular}{ccccc}
       & 0yr & 1yr & 2yr & 3yr \\\hline
      $C$ & 50 & 55 & 57.5 & 58.75 \\
      $D$ & 50 & 45 & 42.5 & 41.25
    \end{tabular}
  \end{center}
\item How successful was your guess? It appears as though the population divide is heading towards 60\% for city $C$ and 40\% for city $D$.
\item It turns out that the dynamics of this scenario are controled by a matrix, and the asymptotic population distribution is an \emph{eigenvector} that we associate with the matrix.
\item Given a square matrix $A$, a vector $\mathbf{x}$ is called an \emph{eigenvector} of $A$ if $A\mathbf{x}=\lambda\mathbf{x}$ for some real or complex number $\lambda$. The number lambda is called an \emph{eigenvalue} of $A$.
\item Eigenvectors are special because the matrix doesn't change what direction they point in. It just scales them and no more.
\item Here is the theoretical method we will use to find eigenvalues and eigenvectors of a matrix.
  \begin{itemize}
  \item Step 0. Reconsider the eigenvalue/eigenvector equation as $A\mathbf{x}-\lambda\mathbf{x}=0$ or $(A-\lambda I)\mathbf{x}=\mathbf{0}$
  \item Step 1. By the above rewriting, we wish to find out when $A-\lambda I$ has a nontrivial null space. We know this happens when $A-\lambda I$ is singular, so set $\det(A-\lambda I)=0$ and solve for $\lambda$.
  \item Step 2. For each solution $\lambda$ obtained in step $1$, plug it into $A-\lambda I$ and calculate the null space basis. This is the corresponding eigenvector $\mathbf{x}$. (Actually there may be several corresponding eigenvectors.)
  \end{itemize}
\item Example.
  \[A=\begin{bmatrix}1&2\\2&4\end{bmatrix}
  \]
  We find that it has eigenvalues $0$ and $5$, with corresponding eigenvectors $(-2,1)$ and $(1,2)$. Check!
\item For our next example let's return to the case of cities $C$ and $D$ as above. The matrix that carries each year's population vector to the next year's population vector is
  \[A=\begin{bmatrix}.8&.3\\.2&.7\end{bmatrix}
  \]
  We find that it has eigenvalues $1$ and $0$, with corresponding eigenvectors $(1.5,1)$ and $(1,-1)$. We will see later that the eigenvector corresponding to $\lambda=1$ determines the asymptotic population distribution (renormalize it!).
\item Example.
  \[A=\begin{bmatrix}1&-1&0\\-1&2&-1\\0&-1&1\end{bmatrix}
  \]
  We find that it has eigenvalues $0,1,3$ with corresponding eigenvectors $(1,1,1)$, $(-1,0,1)$, $(1,-2,1)$. Check!
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Diagonalization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 6.2.

\begin{itemize}
\item For an $n\times n$ matrix $A$, we have said that $\lambda$ is an eigenvalue and $\mathbf{x}$ is an eigenvector of $A$ when the equation holds: $A\mathbf{x}=\lambda\mathbf{x}$.
\item In most cases, there are $n$ linearly independent eigenvectors---thus they constitute a \emph{basis} for $\RR^n$. We will see that $A$ acts really nicely in this basis, in fact, diagonally.
\item To see what this means, assuming $A$ has $n$ eigenvectors, we can form the $n\times n$ matrix $S$ whose columns are the eigenvectors.
\item What is $AS$? For each column $\mathbf{x}$ of $S$, you get a column $\lambda\mathbf{x}$ of $AS$. [Do example]
\item In symbols, we have
  \[A\begin{bmatrix}
      \vdots&&\vdots\\
      \mathbf{x}_1&\cdots&\mathbf{x}_n\\
      \vdots&&\vdots
    \end{bmatrix}
    =\begin{bmatrix}
      \vdots&&\vdots\\
      \lambda_1\mathbf{x}_1&\cdots&\lambda_n\mathbf{x}_n\\
      \vdots&&\vdots
    \end{bmatrix}
    =\begin{bmatrix}
      \vdots&&\vdots\\
      \mathbf{x}_1&\cdots&\mathbf{x}_n\\
      \vdots&&\vdots
    \end{bmatrix}
    \begin{bmatrix}
      \lambda_1\\&\ddots&\\&&\lambda_n
    \end{bmatrix}
  \]
  \item Thus if $S$ is the matrix of eigenvectors, and $\Lambda$ is the diagonal matrix of eigenvalues (in the matching order), we conclude that
  \[AS=S\Lambda
  \]
  and $A$ may be decomposed as:
  \[A=S\Lambda S^{-1}
  \]
\item Thus while most matrices $A$ are not diagonal, it is possible to transform them so that they are!
\item Example. Diagonalize the matrix by writing $A=S\Lambda S^{-1}$.
  \[A=\begin{bmatrix}1&5\\&6\end{bmatrix}
  \]
  the solution is
  \[A=\begin{bmatrix}1&1\\&1\end{bmatrix}
    \begin{bmatrix}1&\\&6\end{bmatrix}
    \begin{bmatrix}1&-1\\&1\end{bmatrix}
  \]
\item There are many applications of diagonalization, but one is the calculation of matrix powers. If you can diagonalize $A$ as $A=S\Lambda S^{-1}$, then you can also diagonalize its powers as $A^n=S\Lambda^nS^{-1}$. (Show.)
\item We will see this come in handy when we want to study the city population problem after $n$ years or in the limit.
\item Example. For $A$ in the previous example, find $A^{100}$. The answer is
  \[A^{100}=\begin{bmatrix}1&1\\&1\end{bmatrix}
    \begin{bmatrix}1&\\&6\end{bmatrix}^{100}
    \begin{bmatrix}1&-1\\&1\end{bmatrix}
    =\begin{bmatrix}1&-1+6^{100}\\&6^{100}\end{bmatrix}
  \]
\item Fibonacci application. Recall that the Fibonacci sequence begins with $F_0=0$, $F_1=1$ and is defined recursively by $F_{n+1}=F_{n+1}+F_n$. Find an absolute formula for $F_n$.
  \begin{itemize}
  \item Step 1. Using the variables $F_{n+1}$ and $F_n$, the process is governed by the equations
    \[\systeme{F_{n+1}=F_n+F_{n-1},F_n=F_n}
    \]
    Or in matrix terms we have
    \[\begin{bmatrix}F_{n+1}\\F_n\end{bmatrix}
      =\begin{bmatrix}1&1\\1&0\end{bmatrix}
      \begin{bmatrix}F_n\\F_{n-1}\end{bmatrix}
    \]
    Denoting this square matrix by $A$, we want to find
    \[\begin{bmatrix}F_{n+1}\\F_n\end{bmatrix}
      =A^n\begin{bmatrix}1\\0\end{bmatrix}
    \]
  \item Step 2. We diagonalize the matrix $A$ above. We find that the two eigenvalues are $\lambda_1=(1+\sqrt{5})/2$ and $\lambda_2=(1-\sqrt{5})/2$. The corresponding eigenvectors are $(\lambda_1,1)$ and $(\lambda_2,1)$. Thus we can diagonalize $A$ as
    \[A=\begin{bmatrix}\lambda_1&\lambda_2\\1&1\end{bmatrix}
      \begin{bmatrix}\lambda_1\\&\lambda_2\end{bmatrix}
      \frac{1}{\sqrt{5}}
      \begin{bmatrix}1&1/\lambda_1\\-1&\lambda_1\end{bmatrix}
      \]
    \item Step 3. We are really interested in
      \begin{align*}
        \begin{bmatrix}F_{n+1}\\F_n\end{bmatrix}
        &=A^n\begin{bmatrix}1\\0\end{bmatrix}\\
        &=\begin{bmatrix}\lambda_1&\lambda_2\\1&1\end{bmatrix}
          \begin{bmatrix}\lambda_1\\&\lambda_2\end{bmatrix}^n
          \frac{1}{\sqrt{5}}
          \begin{bmatrix}1&-\lambda_2\\
            -1&\lambda_1\end{bmatrix}
          \begin{bmatrix}1\\0\end{bmatrix}\\
        &=\frac{1}{\sqrt{5}}
          \begin{bmatrix}\lambda_1&
            \lambda_2\\1&1\end{bmatrix}
          \begin{bmatrix}\lambda_1^n\\
            &(\lambda_2)^n\end{bmatrix}
          \begin{bmatrix}1&-\lambda_2\\
            -1&\lambda_1\end{bmatrix}
          \begin{bmatrix}1\\0\end{bmatrix}\\
        &=\frac{1}{\sqrt{5}}
          \begin{bmatrix}
            \lambda_1^{n+1}-(\lambda_2)^{n+1}\\
            \lambda_1^n-(\lambda_2)^n
          \end{bmatrix}
      \end{align*}
      We can finally conclude the formula
      \[F_n=\frac{1}{\sqrt{5}}
        \left(\left(\frac{1+\sqrt{5}}{2}\right)^n
        -\left(\frac{1-\sqrt{5}}{2}\right)^n\right)
      \]
    \end{itemize}
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Markov matrices}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 10.3.

\begin{itemize}
\item Armed with diagonalization, we now return to the city population problem we initially posed:
\item Every year 20\% of the population of city $C$ moves to city $D$, and 30\% of city $D$ moves to city $C$. What is the eventual population distribution?
\item The process is controlled by the matrix
  \[A=\begin{bmatrix}.8&.3\\.2&.7\end{bmatrix}
  \]
  The population distribution in the $n$th year is given by
  \[\begin{bmatrix}C_n\\D_n\end{bmatrix}
    =A^n\begin{bmatrix}C_0\\D_0\end{bmatrix}
  \]
\item To find $A^n$ we need to diagonalize. We calculated that the eigenvalues or $\lambda=1,.5$ and the corresponding eigenvectors are $(1.5,1)$ and $(-1,1)$. Thus we have
  \begin{align*}
    A^n\begin{bmatrix}C_0\\D_0\end{bmatrix}
    &=\begin{bmatrix}1.5&-1\\1&1\end{bmatrix}
      \begin{bmatrix}1\\&.5^n\end{bmatrix}
      \frac{1}{2.5}
      \begin{bmatrix}1&1\\-1&1.5\end{bmatrix}
  \end{align*}
  In the limit this leads to
  \begin{align*}
    A^\infty\begin{bmatrix}C_0\\D_0\end{bmatrix}
    &=\frac{1}{2.5}\begin{bmatrix}1.5&1.5\\1&1\end{bmatrix}
       \begin{bmatrix}C_0\\D_0\end{bmatrix}\\
    &=\frac{1}{2.5}\begin{bmatrix}1.5\\1\end{bmatrix}
     (C_0+D_0)
  \end{align*}
\item The calculation will essentially always turn out this way. A vector that sums to $1$ is called a \emph{probability vector}. A matrix whose columns are all probability vectors is called a \emph{Markov matrix}.
\item Theorem. Suppose $A$ is a Markov matrix with nonzero entries and $\mathbf{x}$ is a probability vector. Then $\lim_{n\to\infty}A^n\mathbf{x}$ is equal to the normalized eigenvector corresponding to $\lambda=1$.
\item This theorem about Markov matrices is a special case of the much more powerful Perron-Frobenius theorem: If $A$ is a positive matrix, then its dominant (greatest) eigenvalue $\lambda_m$ is real and positive. Then the growth rate of the powers $A^n\mathbf{x}$ is given by $\lambda_m$ and the limiting distribution is given by the corresponding eigenvector $\mathbf{x}_m$.
\item Example. A Leslie matrix describes how a population changes through the generations.
\item For instance suppose you observe the fertility rates and survival rates of three generations: Group $1$ is age $0$ to $20$, Group $2$ is age $20$ to $40$, and Group $3$ is age $40$ to $60$. The fertility rates $F_n$ are the number of children each female in Group $n$ has in $20$ years, the survival rates $S_n$ are the proportion of Group $n$ that survives to group $n+1$.
\item Suppose you measure $F_1=.04$, $F_2=1.1$, $F_3=.01$, $S_1=.98$, and $S_2=.92$. Then the process is controlled by the matrix
  \[A=\begin{bmatrix}.04&1.1&.01\\.98\\&.92\end{bmatrix}
  \]
  This isn't a Markov matrix, but it is still subject to the Perron--Frobenius theorem.
\item Using a calculator I found that the largest real eigenvalue was $\lambda=1.06$. Thus the population grows at a rate of 6\% per generation. The corresponding eigenvector, renormalized to sum to $1$, is $(.37,.34,.29)$, and tells the asymptotic distribution of the generations.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Systems of differential equations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S 6.3.

\begin{itemize}
\item In our study of discrete processes such as Markov processes, we always have the recursive equation $\mathbf{u}_{n+1}=A\mathbf{u}_n$ and an initial state $\mathbf{u}_0$ given. Hear the variables at time $n+1$ depend in a linear way on the variables at time $n$.
\item The continuous analog of this situation is a differential equation $\mathbf{u}'=A\mathbf{u}$ and an initial condition $\mathbf{u}(0)=\mathbf{u}_0$. Here the growth rate of the variables depends in a linear way on the variables themselves.
\item For a simple one-variable example, unrestricted population growth is the situation when $y'=ky$, $y(0)=y_0$. We know that the solutions to this differential equation is $y=y_0e^{kt}$, that is, exponential growth.
\item In the multivariable version, we simply replace a single scalar $k$ with a matrix $A$. We obtain a linear system of first-order ordinary differential equations.
  \[\begin{cases}
      \mathbf{u}'=A\mathbf{u}\\
      \mathbf{u}(0)=\mathbf{u}_0
    \end{cases}
  \]
\item It turns out that we can write the solution to such an equation as
  \[\mathbf{u}(t)=e^{At}\mathbf{u}_0
  \]
  But this leaves us with three burning questions. First, what exactly is $e^{At}$? Second, why does it solve the linear system of differential equations? And third, how can we calculate it?
\item For the first question, recall that $e^x$ may be defined using the series $e^x=1+x+x^2/2+x^3/3!+\cdots$. Thus we define
  \[e^{At}=I+At+\frac{(At)^2}{2}+\frac{(At)^3}{3!}+\cdots
  \]
  It is a fact that this sum always converges.
\item For the second question, we simply take the derivative of $\mathbf{u}=e^{At}\mathbf{u}_0$:
  \begin{align*}
    \frac{d}{dt}e^{At}\mathbf{u}_0
    &=\frac{d}{dt}\left[
      I+At+\frac{(At)^2}{2}+\frac{(At)^3}{3!}+\cdots
      \right]\mathbf{u}_0\\
    &=\left[0+A+A^2t+A\frac{(At)^2}{2}+\cdots\right]
      \mathbf{u}_0\\
    &=A\left[I+At+\frac{(At)^2}{2}+\cdots\right]\mathbf{u}_0\\
    &=Ae^{At}\mathbf{u}_0
  \end{align*}
\item Finally, if $A$ is diagonalizable then one can compute $e^{At}$ using diagonalization. To see this, if $A=S\Lambda S^{-1}$ then
  \begin{align*}
    e^A&=e^{S\Lambda S^{-1}}\\
       &=I+S\Lambda S^{-1}t+\frac{S\Lambda^2S^{-1}t^2}{2}
         +\cdots\\
       &=S\left[I+\Lambda t+\frac{\Lambda^2t^2}{2}+\cdots
         \right]S^{-1}\\
       &=Se^{\Lambda t}S^{-1}
  \end{align*}
  Since $\Lambda$ is diagonal it is easy to calculate $e^{\Lambda t}$ by just applying $e^{at}$ to each of its diagonal entries.
\item Ok, that was the theory, now comes the action. The diffusion of a gas between two adjacent chambers satisfies the equations:
    \[\begin{cases}
      u'=-u+v & u(0)=15\\
      v'=u-v & v(0)=5
    \end{cases}
  \]
  To solve the system we use the matrix
  \[A=\begin{bmatrix}-1&1\\1&-1\end{bmatrix}
  \]
  Which has the diagonalization
  \[A=\begin{bmatrix}1&1\\1&-1\end{bmatrix}
    \begin{bmatrix}0\\&-2\end{bmatrix}
    \frac{1}{-2}
    \begin{bmatrix}-1&-1\\-1&1\end{bmatrix}
  \]
  It follows that
  \[e^{At}=\begin{bmatrix}1&1\\1&-1\end{bmatrix}
    \begin{bmatrix}e^{0t}\\&e^{-2t}\end{bmatrix}
    \frac{1}{-2}
    \begin{bmatrix}-1&-1\\-1&1\end{bmatrix}
  \]
  And finally
  \[\mathbf{u}=e^{At}\mathbf{u}_0
    =\begin{bmatrix}1&1\\1&-1\end{bmatrix}
    \begin{bmatrix}1\\&e^{-2t}\end{bmatrix}
    \frac{1}{-2}
    \begin{bmatrix}-1&-1\\-1&1\end{bmatrix}
    \begin{bmatrix}15\\5\end{bmatrix}
  \]
  Calculating we arrive at the solution
  \[\mathbf{u}
    =\begin{bmatrix}10+5e^{-2t}\\10-5e^{-2t}\end{bmatrix}
  \]
  We observe that in the limit, the two chambers will have an equal proportion of gas.
\end{itemize}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Markov processes and search rankings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Based on ILA 5th, \S ??

\begin{itemize}
\item We conclude with one final application of Markov matrices: web search.
\item When you perform a web search, two things happen. First, the system filters just the sites matching your search terms. Second, the system ranks the sites according to popularity. The second part is handled by an interesting application of linear algebra which google calls Pagerank.
\item To see how it works, imagine a very small internet with just five pages. We depict the pages as nodes and the links between them as arrows.
  \begin{center}
    \begin{tikzpicture}[scale=1.2]
      \node at (0:1) (A) {$A$};
      \node at (72:1) (B) {$B$}
      edge[->] (A) edge[<-,bend left] (A);
      \node at (144:1) (C) {$C$} edge[<-] (A) edge[<-,bend left] (B);
      \node at (216:1) (D) {$D$} edge[<-] (A) edge[<-,bend left] (C);
      \node at (288:1) (E) {$E$} edge[<-] (C) edge[<-,bend left] (D);
    \end{tikzpicture}
  \end{center}
\item We model browsing with a matrix as follows. We postulate a generic user who clicks links on the pages. Each time she opens a page, she chooses one of the links at random, with equal probability. If a page has no links, she chooses one of the other pages at random. Thus for this internet we have
  \[A=\begin{bmatrix}
      &1/2&&&1/5\\
      1/3&&&&1/5\\
      1/3&1/2&&&1/5\\
      1/3&&1/2&&1/5\\
      &&1/2&1&1/5
    \end{bmatrix}
  \]
\item If we initially assume that the pages are visited with some probability vector $\mathbf{u}_0$, then $A^n\mathbf{u}_0$ gives the likelihood of being on each page after $n$ clicks.
\item The steady state solution $A^\infty\mathbf{u}_0$ gives us an unbiased measurement of the relative popularity of each web page!
\item Since the matrix is Markov, we can calculate this simply by finding the eigenvector corresponding to $\lambda=1$. Using Octave, we find that $A-I$ has the RREF form
  \[A-I\sim\begin{bmatrix}
      1&&&&-9/25\\
      &1&&&-8/25\\
      &&1&&-12/25\\
      &&&1&-14/25\\
      \phantom{1}
    \end{bmatrix}
  \]
  We conclude that the eigenvector is
  \[(9/25,8/25,12/25,14/25,1)
  \]
  Which we can renormalize to
  \[(0.13,0.12,0.18,0.21,0.37)
  \]
  This tells us that E is the most popular, followed by D.
\end{itemize}

\end{document}
